\begin{abstract}
\begin{abstract}
Recent studies highlight a common problem in clinical machine learning: models that perform well during development often do not generalize to new patient groups. This problem is seen across areas such as psychiatry, genomics, and biomarker research. For example, Roy H. Perlis, Chekroud, and colleagues found that the accuracy of schizophrenia treatment models dropped from 0.72 during internal testing to 0.54 on separate datasets~\cite{chekroud2024illusory}. This drop in performance is usually linked to issues such as changes in data, misleading connections, hidden factors, and evaluation errors~\cite{whalen2022pitfalls, bendavid2010theory}.

Although new methods like invariant risk minimization and causal modeling offer theoretical fixes, they are rarely used in practice~\cite{arjovsky2019irm, pavlovic2024improving}. These methods usually require strong assumptions, such as data from multiple training settings or detailed cause-and-effect models, which are rarely available in real clinical settings. Because of this, experts are good at explaining why models fail after the fact but still lack useful tools to predict these problems before using the models.

To fill this gap, a practical three-step process for checking risks before deployment is suggested. First, differences between the training data and the target patient groups are measured using tools such as the Maximum Mean Discrepancy, Kullback--Leibler divergence, and Wasserstein distance~\cite{gretton2012kernel, duchi2018learning, rahimian2019dro}. Second, the stability of individual features is tested across different settings using cross-context permutation importance to find unreliable predictors. Third, these results guide clear decisions about deployment, such as whether to deploy, adjust, limit, or hold back the model, supported by data-based limits and ongoing monitoring after deployment.

The objective is to deliver a practical, accessible pipeline that clinical machine learning teams can implement using standard tools, without specialized expertise in causal inference or complex modeling assumptions.
\end{abstract}
\end{abstract}