\section{Research Gap and Objectives}

The literature points to a clear imbalance. Researchers have built strong tools for understanding why models fail  shift theory (Ben-David et al., 2010; Wörheide et al., 2021), pitfall checklists (Whalen et al., 2022), real-world evidence of collapse (Chekroud et al., 2024; Springer \& Doering, 2021), reporting standards (Collins et al., 2024), and causal methods (Pavlovic et al., 2024; Arjovsky et al., 2019). But none of these tools can tell a practitioner, before deployment, whether their model will work in a new setting. Chekroud et al. show the problem but test no solutions. Pavlovic et al. offer the best solution but it requires a causal graph most teams cannot build. Wörheide et al. test corrections against only one type of shift and recommend DRO without experimental evidence. Collins et al. require discussion of generalisation limits but set no standard for what that discussion should include. Nobody has combined distributional measurement, feature-level analysis, and deployment guidance into one usable workflow. The gap is not that the field does not know about the problem  every paper acknowledges it  but that nobody has assembled the pieces into something that works in practice.\\
Research Question: To what extent can pre-deployment distributional diagnostics  combining quantitative measures of dataset shift with feature-level stability analysis  predict how much a clinical prediction model's performance will drop when applied to a new population, and what thresholds can reliably tell apart settings where a model is safe to deploy from settings where it is not?\\
Objectives:\\
Objective 1 (Week 1). Implement three distributional divergence measures  MMD, KL divergence, and Wasserstein distance  and apply them to publicly available multi-context clinical data where performance drops across contexts are already known. For each pair of contexts, compute the divergence score and measure how well it correlates with the actual drop in balanced accuracy. Deliverable: a ranked comparison showing which metric is most useful for predicting real performance loss, with correlation values and confidence intervals.\\
Objective 2 (Week 2). Build a feature stability scoring method that labels each predictor in a model as either shift-stable (it stays important regardless of which context the model is tested on) or shift-sensitive (its importance changes depending on context). Do this using a leave-one-context-out approach with permutation importance, and check whether the labels match reality by comparing against actual cross-context performance. Target: at least 70\% agreement between predicted and observed instability. Deliverable: a working tool that takes any multi-context dataset and produces a per-feature stability report.\\
Objective 3 (Week 3). Combine the divergence scores from Objective 1 and the stability profiles from Objective 2 into a decision flowchart with four possible outputs: deploy with standard monitoring, deploy with extra monitoring, retrain using only stable features, or do not deploy. Set the thresholds using the empirical results from Weeks 1 and 2, and include a simple monitoring plan for after deployment  what to measure, how often, and what should trigger a review. Test the framework on at least one held-out context that was not used to set the thresholds, and be honest about where it works and where it does not. Deliverable: a documented decision protocol with a flowchart, threshold values, and a clear limitations section.